A White Noise Approach to Evolutionary Ecology

### 2020-04-28 Current Version/Current-Version.Rmd
reorganize

```{r global_options, include=FALSE}
#knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

@@

\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}

where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We also assume $(1+h)C\leq m(h,x)$ for some constant $C>0$ so that $m(h,x)$ is linear or more than linear in $h$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}

with the initial condition $\nu(x,0)$ described above. Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies

@@

\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation}

where $r\in\mathbb{R}$ is the intrinsic rate of growth in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and $\kappa(x-y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. In \S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is closely related to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

\begin{multline}\label{detmean}
\frac{d}{dt}\bar x(t)=\frac{d}{dt}\int_{-\infty}^{+\infty} x p(x,t)dx=\int_{-\infty}^{+\infty} x\frac{\partial}{\partial t} p(x,t)dx \\
=\int_{-\infty}^{+\infty} x\big(m(\nu,x)-\bar m(t)\big) p(x,t)+x\frac{\mu}{2}\frac{\partial^2}{\partial x^2} p(x,t)dx
=\mathrm{Cov}_t\Big(m(\nu,x),x\Big).
\end{multline}

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/wn.png",wn,width=5,height=2.5)
```

@@

\begin{equation}
\|f\|_t=\sqrt{\mathbb{E}\left(\int_0^t\int_\mathbb{R}f^2(x,s)dxds\right)},
\end{equation}

and make use of the convention $f=g$ if $\|f-g\|_t=0$ for all $t\geq0$. Later, when we investigate stochastic dynamics of the abundance density $\nu(x,t)$, we will find $\sqrt\nu\in\mathscr{N}_2$. This enables us to utilize the heuristics developed in this section for the derivation of SDE describing the stochastic dynamics of $N(t)$, $\bar x(t)$ and $\sigma^2(t)$.

@@

\begin{equation}\label{exp_WN}
\mathbb{E}\big(\mathbf W_t(f)\big)=\mathbb{E}\big(\mathbf W_t(g)\big)=0,
\end{equation}
\begin{equation}\label{cov_WN}
\mathbb{C}\big(\mathbf W_{t_1}(f),\mathbf W_{t_2}(g)\big)=\mathbb{E}\left(\int_0^{t_1\wedge t_2}\int_\mathbb{R} f(x,s)g(x,s)dxds\right),
\end{equation}

where $t_1\wedge t_2=\min(t_1,t_2)$ and $\mathbb{C}$ denotes covariance with respect to the underlying probability space. In particular, denoting $\mathbb{V}$ the variance operator with respect to the underlying probability space, we have $\mathbb{V}\big(\mathbf W_t(f)\big)=\|f\|_t^2$ for all $t\geq0$ and $f\in\mathscr{N}_2$. The operators $\mathbb{E}$ and $\mathbb{C}$ are to be distinguished from $\bar f(t)$ and $\mathrm{Cov}_t(f,g)$ which denote expectation and covariance with respect to phenotypic diversity at time $t\geq0$.

@@

\begin{equation}
\|I_D\|_t^2=\mathbb{E}\left(\int_0^t\int_\mathbb{R}I_D^2(x,s)dxds\right)=t\int_Ddx=t|D|<+\infty, 
\end{equation}

where $|D|$ denotes the length of $D$, we have $I_D\in\mathscr{N}_2$. Thus, using equations (\ref{exp_WN}) and (\ref{cov_WN}) and adopting the informal notation introduced in equation (\ref{informal}), we can write the following

@@

\begin{equation}
\hat f(x,t)=\frac{f(x,t)}{\sqrt{\int_\mathbb{R}f^2(y,t)dy}} \ \text{ and } \ 
d\hat{\mathbf W}_t(f)=\left(\int_\mathbb{R}\hat f(x,t)\dot W(x,t)dx\right)dt
\end{equation}

so that

@@

\begin{equation}
\mathbb{E}\left(\int_0^td\hat{\mathbf W}_s(f)\right)=0, \ \mathbb{C}\left(\int_0^{t_1}d\hat{\mathbf W}_s(f),\int_0^{t_2}d\hat{\mathbf W}_s(f)\right)=t_1\wedge t_2
\end{equation}

and in particular, as a function of $t$, $\int_0^td\hat{\mathbf W}_s(f)$ is a standard Brownian motion for any $f\in\mathscr{N}_2$. Hence, $d\hat{\mathbf W}_t(f)$ is analogous to the traditional shorthand used to denote stochastic differentials. Thus, equation (\ref{cov_WN_xi}) effectively extends It&ocirc;'s multiplication table to:

@@

_Branching Brownian motion (BBM):_
A BBM tracks individuals navigating $d$-dimensional Euclidean space that reproduce and senesce between exponentially distributed intervals. Unlike other stochastic processes that take values in $\mathbb{R}^d$, BBM takes values in the set of _non-negative finite measures_ over $\mathbb{R}^d$. Intuitively, one can think of a finite measure as a function that maps subsets of $\mathbb{R}^d$ to real numbers. To be formal, we only consider the Borel subsets of $\mathbb{R}^d$ corresponding to the Euclidean metric, but understanding this technicality is not crucial to our discussion. In particular, denoting $X_t$ a BBM, for a subset $D\subset\mathbb{R}^d$, $X_t(D)$ returns the (random) number of individuals alive within the region $D$ at time $t\geq0$. The BBM has three main components:

@@

\begin{multline}\label{neutral_int}
\int_\mathbb{R}\nu(x,t)f(x)dx-\int_\mathbb{R}\nu(x,0)f(x)dx=\int_0^t\int_\mathbb{R}\nu(x,s)\left(mf(x)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}f(x)\right)dsdx \\
+\int_0^t\int_\mathbb{R}f(x)\sqrt{V\nu(x,s)}\dot W(x,s)dxds, \ \forall \ f\in C_b^2(\mathbb{R}).
\end{multline}

@@

_Interacting superprocesses:_
In the above subsection we reviewed the diffusion limit of an especially tractable measure-valued branching process. However, we found the simplicity of this process restricts us from modelling nonlocal effects on the fitness of individuals. Here, we discuss superprocesses that account for such effects. The existence of diffusion limits for a class of measure-valued branching processes involving nonlocal effects has been treated by Méléard and Roelly [-@meleard1992interacting; -@Mlard1993]. The interactions can manifest as dependencies of the spatial movement or reproductive law of individuals on their position $x$ and the state of the whole population described by $X_t$. An important result of Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] is a theorem that provides sufficient conditions to construct a sequence of rescaled measure-valued branching processes that converge to a generalization of the Dawson-Watanabe superprocess that includes interactions. The rescaling is analogous to that described above for non-interacting Dawson-Watanabe superprocesses, but now the reproductive law described by $\mathscr{W}(X_t,x)$ and $V(X_t,x)$, branching rate $\lambda(X_t,x)$ and diffusion parameter $\sqrt{\mu(X_t,x)}$ are allowed to depend on the whole population $X_t$ and individual location $x$. In Figure \ref{rescaled} we demonstrate this rescaling in discrete time for a population experiencing stabilizing selection and logistic growth. Since time is discretized, the process we simulate is formally a branching random walk. For further details on our simulation see SM \S\ref{numerical}.

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/rescaled_plots.png",rescaled_plots,device="png",width=6,height=6,units="in",dpi=350)
```

@@

\begin{multline}
N(t)=N(0)+\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)\cdot1+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}1\right)+1\sqrt{V\nu(x,s)}\dot W(x,s)dsdx \\
=N(0)+\int_0^t\bar m(s)N(s)dt+\int_0^t\sqrt{VN(s)}d\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)}),
\end{multline}

where

\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}

and

@@

\begin{subequations}\label{gen_eqns}
\begin{equation}\label{N}
dN(t)=\bar m(t)N(t)dt+\sqrt{V N(t)}dW_1(t),
\end{equation}
\begin{equation}\label{xbar_gen}
d\bar x(t)=\mathrm{Cov}_t\Big(x,m(\nu,x)\Big)dt+\sqrt{V\frac{\sigma^2(t)}{N(t)}}dW_2(t),
\end{equation}
\begin{equation}\label{sig2_gen}
d\sigma^2(t)=\mathrm{Cov}_t\Big((x-\bar x(t))^2,m(\nu,x)\Big)dt+\left(\mu-V\frac{\sigma^2(t)}{N(t)}\right)dt+\sqrt{V\frac{\overline{(x-\bar x(t))^4}-\sigma^4(t)}{N(t)}}dW_3(t),
\end{equation}
\end{subequations}

where $W_1$, $W_2$ and $W_3$ are standard Brownian motions. We note that conditions on the growth rate $m$ to guarantee existence and uniqueness of solutions to (\ref{xbar_gen}) and (\ref{sig2_gen}) have yet to be investigated. However, our results on the deterministic PDE suggest that $m(h,x)$ bounded above and differentiable in both arguments is sufficient. Dividing by $dt$ one can interpret equations (\ref{gen_eqns}) as if they are ordinary differential equations, but this not technically rigorous since Brownian motion is nowhere differentiable with respect to time. In SM \S\ref{SDE_DERIV} we show that in general $W_1$ is independent of both $W_2$ and $W_3$, but $W_2$ and $W_3$ covary.

@@

\begin{equation}\label{covxm}
\mathrm{Cov}(x,m)=\sigma^2\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right),
\end{equation}
\begin{equation}
\mathrm{Cov}\Big((x-\bar x)^2,m\Big)=2\sigma^4\left(\frac{\partial\bar m}{\partial\sigma^2}-\overline{\frac{\partial m}{\partial\sigma^2}}\right)
\end{equation}

and $\overline{(x-\bar x)^4}=3\sigma^4$. In particular, these results imply

@@

For species $i$ we inherit the above notation for trait value, distribution, average, variance, abundance, et cetera, except with an $i$ in the subscript. Real world examples of niche axes include the size of seeds consumed competing finch species and the date of activity in a season for pollinators competing for floral resources. For mathematical convenience, we model the axis of resources by the real line $\mathbb{R}$. The value of a resouce along this axis is denoted by the symbol $\zeta$. For an individual in species $i$, we assume the resource utilization curve $u_i$ can be written as

@@

\begin{equation}
\mathcal O_{ij}(x_i-x_j)=\int_{-\infty}^{+\infty}u_i(\zeta,x_i)u_j(\zeta,x_j)d\zeta=\frac{U_iU_j}{\sqrt{2\pi(w_i+w_j)}}\exp\left(-\frac{(x_i-x_j)^2}{2(w_i+w_j)}\right).
\end{equation}

A notable criticism of using niche overlap to measure the intensity of competition points to cases where populations competiting on multiple niche axes exhibit overlap on at least one of the axes, but no overall niche overlap [@Holt1987]. Thus niche overlap on lower-dimensional projections of some multivariate niche space does not imply the populations compete. To illustrate with a simple example, consider two populations competing for space on the plane $\mathbb{R}^2$. If the spatial distributions of the two populations overlap, then they will overlap on both spatial axes. However, if the populations do not overlap on at least one of the spatial axes, they will have no overall spatial overlap. Furthermore, even if the species overlap on both spatial axes, they need not have any overall spatial overlap. This final result corresponds to the fact that components of niche space do not necessarily interact multiplicatively to determine the consequences for the intensity of competition. In another component of Holt's [-@Holt1987] critique, an argument is made for the potential of competition occuring without any overlap in niche space. However, this argument is based on the practical difficulty of identifying every resource axis populations are competing on and how these axes interact to determine fitness consequences. Our model avoids these caveats by assuming competition only occurs along a single dimensional resource gradient.

To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ decreases the expected reproductive output for the individual in species $i$ at the rate $c_i\mathcal{O}_{ij}(x_i-x_j)$ for some $c_i>0$. We refer to $c_i$ as the sensitivity to competition for species $i$. The term $c_i\mathcal{O}_{ij}(x_i-x_j)$ coincides with a special case of a term used to capture competition in Dawson's geostochastic logistic model, an SPDE model developed to study the combined effects of demographic stochasticity, spatial dispersion and locally finite carrying capacity [@Dawson1978]. In relation to the example fitness function discussed in \S\ref{deterministic}, consider $\kappa(x_i-x_j)=\mathcal{O}_{ij}(x_i-x_j)$. In SM \S\ref{diffuse} we combine our treatment of resource competition with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/community_dynamics.png",community_plot,width=6.5,height=5,dpi=320)

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/on_pl.png",on_pl,device="png",width=3.5,height=3.5,units="in",dpi=350)

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/interaction_networks.png",coevs,device="png",width=6,height=6,units="in",dpi=350)

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/.A White Noise Approach To Evolutionary Ecology/on_pl.png",on_pl2,device="png",width=3.5,height=3.5,units="in")
```

@@

>>>>

```{r global_options, include=FALSE}
options(tinytex.clean = FALSE)
#knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

@@

\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We also assume $(1+h)C\leq m(h,x)$ for some constant $C>0$ so that $m(h,x)$ is linear or more than linear in $h$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies

@@

\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation}
where $r\in\mathbb{R}$ is the intrinsic rate of growth in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and $\kappa(x-y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. In \S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is closely related to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.

@@

\begin{multline}\label{detmean}
\frac{d}{dt}\bar x(t)=\frac{d}{dt}\int_{-\infty}^{+\infty} x p(x,t)dx=\int_{-\infty}^{+\infty} x\frac{\partial}{\partial t} p(x,t)dx \\
=\int_{-\infty}^{+\infty} x\big(m(\nu,x)-\bar m(t)\big) p(x,t)+x\frac{\mu}{2}\frac{\partial^2}{\partial x^2} p(x,t)dx \\
=\mathrm{Cov}_t\Big(m(\nu,x),x\Big).
\end{multline}

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/Current Version/wn.png",wn,width=5,height=2.5)
```

@@

\begin{equation}
\|f\|_t=\sqrt{\mathbb{E}\left(\int_0^t\int_\mathbb{R}f^2(x,s)dxds\right)},
\end{equation}
and make use of the convention $f=g$ if $\|f-g\|_t=0$ for all $t\geq0$. Later, when we investigate stochastic dynamics of the abundance density $\nu(x,t)$, we will find $\sqrt\nu\in\mathscr{N}_2$. This enables us to utilize the heuristics developed in this section for the derivation of SDE describing the stochastic dynamics of $N(t)$, $\bar x(t)$ and $\sigma^2(t)$.

@@

\begin{equation}\label{exp_WN}
\mathbb{E}\big(\mathbf W_t(f)\big)=\mathbb{E}\big(\mathbf W_t(g)\big)=0,
\end{equation}
\begin{equation}\label{cov_WN}
\mathbb{C}\big(\mathbf W_{t_1}(f),\mathbf W_{t_2}(g)\big)=\mathbb{E}\left(\int_0^{t_1\wedge t_2}\int_\mathbb{R} f(x,s)g(x,s)dxds\right),
\end{equation}
where $t_1\wedge t_2=\min(t_1,t_2)$ and $\mathbb{C}$ denotes covariance with respect to the underlying probability space. In particular, denoting $\mathbb{V}$ the variance operator with respect to the underlying probability space, we have $\mathbb{V}\big(\mathbf W_t(f)\big)=\|f\|_t^2$ for all $t\geq0$ and $f\in\mathscr{N}_2$. The operators $\mathbb{E}$ and $\mathbb{C}$ are to be distinguished from $\bar f(t)$ and $\mathrm{Cov}_t(f,g)$ which denote expectation and covariance with respect to phenotypic diversity at time $t\geq0$.

@@

\begin{equation}
\|I_D\|_t^2=\mathbb{E}\left(\int_0^t\int_\mathbb{R}I_D^2(x,s)dxds\right)=t\int_Ddx=t|D|<+\infty, 
\end{equation}
where $|D|$ denotes the length of $D$, we have $I_D\in\mathscr{N}_2$. Thus, using equations (\ref{exp_WN}) and (\ref{cov_WN}) and adopting the informal notation introduced in equation (\ref{informal}), we can write the following

@@

\begin{equation}
\hat f(x,t)=\frac{f(x,t)}{\sqrt{\int_\mathbb{R}f^2(y,t)dy}} \ \text{ and } \ 
d\hat{\mathbf W}_t(f)=\left(\int_\mathbb{R}\hat f(x,t)\dot W(x,t)dx\right)dt
\end{equation}
so that

@@

\begin{equation}
\mathbb{E}\left(\int_0^td\hat{\mathbf W}_s(f)\right)=0, \ \mathbb{C}\left(\int_0^{t_1}d\hat{\mathbf W}_s(f),\int_0^{t_2}d\hat{\mathbf W}_s(f)\right)=t_1\wedge t_2
\end{equation}
and in particular, as a function of $t$, $\int_0^td\hat{\mathbf W}_s(f)$ is a standard Brownian motion for any $f\in\mathscr{N}_2$. Hence, $d\hat{\mathbf W}_t(f)$ is analogous to the traditional shorthand used to denote stochastic differentials. Thus, equation (\ref{cov_WN_xi}) effectively extends It&ocirc;'s multiplication table to:

@@

\paragraph{Branching Brownian motion (BBM)}
A BBM tracks individuals navigating $d$-dimensional Euclidean space that reproduce and senesce between exponentially distributed intervals. Unlike other stochastic processes that take values in $\mathbb{R}^d$, BBM takes values in the set of _non-negative finite measures_ over $\mathbb{R}^d$. Intuitively, one can think of a finite measure as a function that maps subsets of $\mathbb{R}^d$ to real numbers. To be formal, we only consider the Borel subsets of $\mathbb{R}^d$ corresponding to the Euclidean metric, but understanding this technicality is not crucial to our discussion. In particular, denoting $X_t$ a BBM, for a subset $D\subset\mathbb{R}^d$, $X_t(D)$ returns the (random) number of individuals alive within the region $D$ at time $t\geq0$. The BBM has three main components:

@@

\begin{multline}\label{neutral_int}
\int_\mathbb{R}\nu(x,t)f(x)dx-\int_\mathbb{R}\nu(x,0)f(x)dx\\=\int_0^t\int_\mathbb{R}\nu(x,s)\left(mf(x)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}f(x)\right)dsdx \\
+\int_0^t\int_\mathbb{R}f(x)\sqrt{V\nu(x,s)}\dot W(x,s)dxds, \ \forall \ f\in C_b^2(\mathbb{R}).
\end{multline}

@@

\paragraph{Interacting superprocesses}
In the above subsection we reviewed the diffusion limit of an especially tractable measure-valued branching process. However, we found the simplicity of this process restricts us from modelling nonlocal effects on the fitness of individuals. Here, we discuss superprocesses that account for such effects. The existence of diffusion limits for a class of measure-valued branching processes involving nonlocal effects has been treated by Méléard and Roelly [-@meleard1992interacting; -@Mlard1993]. The interactions can manifest as dependencies of the spatial movement or reproductive law of individuals on their position $x$ and the state of the whole population described by $X_t$. An important result of Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] is a theorem that provides sufficient conditions to construct a sequence of rescaled measure-valued branching processes that converge to a generalization of the Dawson-Watanabe superprocess that includes interactions. The rescaling is analogous to that described above for non-interacting Dawson-Watanabe superprocesses, but now the reproductive law described by $\mathscr{W}(X_t,x)$ and $V(X_t,x)$, branching rate $\lambda(X_t,x)$ and diffusion parameter $\sqrt{\mu(X_t,x)}$ are allowed to depend on the whole population $X_t$ and individual location $x$. In Figure \ref{rescaled} we demonstrate this rescaling in discrete time for a population experiencing stabilizing selection and logistic growth. Since time is discretized, the process we simulate is formally a branching random walk. For further details on our simulation see SM \S\ref{numerical}.

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/Current Version/rescaled_plots.png",rescaled_plots,device="png",width=6,height=6,units="in",dpi=350)
```

@@

\begin{multline}
N(t)-N(0)=\\
\int_0^t\int_\mathbb{R}\nu(x,s)\left(m(\nu,x)\cdot1+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}1\right)+1\sqrt{V\nu(x,s)}\dot W(x,s)dsdx \\
=\int_0^t\bar m(s)N(s)dt+\int_0^t\sqrt{VN(s)}d\hat{\mathbf{W}}_s(\sqrt{\nu(x,s)}),
\end{multline}
where

\begin{equation}
\bar m(t)=\frac{1}{N(t)}\int_\mathbb{R}m(\nu,x)\nu(x,t)dx,
\end{equation}
and

@@

\begin{subequations}\label{gen_eqns}
\begin{equation}\label{N}
dN(t)=\bar m(t)N(t)dt+\sqrt{V N(t)}dW_1(t),
\end{equation}
\begin{equation}\label{xbar_gen}
d\bar x(t)=\mathrm{Cov}_t\Big(x,m(\nu,x)\Big)dt+\sqrt{V\frac{\sigma^2(t)}{N(t)}}dW_2(t),
\end{equation}
\begin{multline}\label{sig2_gen}
d\sigma^2(t)=\mathrm{Cov}_t\Big((x-\bar x(t))^2,m(\nu,x)\Big)dt+\left(\mu-V\frac{\sigma^2(t)}{N(t)}\right)dt\\+\sqrt{V\frac{\overline{(x-\bar x(t))^4}-\sigma^4(t)}{N(t)}}dW_3(t),
\end{multline}
\end{subequations}
where $W_1$, $W_2$ and $W_3$ are standard Brownian motions. We note that conditions on the growth rate $m$ to guarantee existence and uniqueness of solutions to (\ref{xbar_gen}) and (\ref{sig2_gen}) have yet to be investigated. However, our results on the deterministic PDE suggest that $m(h,x)$ bounded above and differentiable in both arguments is sufficient. Dividing by $dt$ one can interpret equations (\ref{gen_eqns}) as if they are ordinary differential equations, but this not technically rigorous since Brownian motion is nowhere differentiable with respect to time. In SM \S\ref{SDE_DERIV} we show that in general $W_1$ is independent of both $W_2$ and $W_3$, but $W_2$ and $W_3$ covary.

@@

\begin{equation}\label{covxm}
\mathrm{Cov}(x,m)=\sigma^2\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right),
\end{equation}
\begin{equation}
\mathrm{Cov}\Big((x-\bar x)^2,m\Big)=2\sigma^4\left(\frac{\partial\bar m}{\partial\sigma^2}-\overline{\frac{\partial m}{\partial\sigma^2}}\right)
\end{equation}
and $\overline{(x-\bar x)^4}=3\sigma^4$. In particular, these results imply

@@

\paragraph{Description of abiotic selection and competition}
For species $i$ we inherit the above notation for trait value, distribution, average, variance, abundance, et cetera, except with an $i$ in the subscript. Real world examples of niche axes include the size of seeds consumed competing finch species and the date of activity in a season for pollinators competing for floral resources. For mathematical convenience, we model the axis of resources by the real line $\mathbb{R}$. The value of a resouce along this axis is denoted by the symbol $\zeta$. For an individual in species $i$, we assume the resource utilization curve $u_i$ can be written as

@@

\begin{multline}
\mathcal O_{ij}(x_i-x_j)=\int_{-\infty}^{+\infty}u_i(\zeta,x_i)u_j(\zeta,x_j)d\zeta\\=\frac{U_iU_j}{\sqrt{2\pi(w_i+w_j)}}\exp\left(-\frac{(x_i-x_j)^2}{2(w_i+w_j)}\right).
\end{multline}

\paragraph{Avoiding caveats of niche theory}
A notable criticism of using niche overlap to measure the intensity of competition points to cases where populations competiting on multiple niche axes exhibit overlap on at least one of the axes, but no overall niche overlap [@Holt1987]. Thus niche overlap on lower-dimensional projections of some multivariate niche space does not imply the populations compete. To illustrate with a simple example, consider two populations competing for space on the plane $\mathbb{R}^2$. If the spatial distributions of the two populations overlap, then they will overlap on both spatial axes. However, if the populations do not overlap on at least one of the spatial axes, they will have no overall spatial overlap. Furthermore, even if the species overlap on both spatial axes, they need not have any overall spatial overlap. This final result corresponds to the fact that components of niche space do not necessarily interact multiplicatively to determine the consequences for the intensity of competition. In another component of Holt's [-@Holt1987] critique, an argument is made for the potential of competition occuring without any overlap in niche space. However, this argument is based on the practical difficulty of identifying every resource axis populations are competing on and how these axes interact to determine fitness consequences. Our model avoids these caveats by assuming competition only occurs along a single dimensional resource gradient.

\paragraph{The fitness function}
To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ decreases the expected reproductive output for the individual in species $i$ at the rate $c_i\mathcal{O}_{ij}(x_i-x_j)$ for some $c_i>0$. We refer to $c_i$ as the sensitivity to competition for species $i$. The term $c_i\mathcal{O}_{ij}(x_i-x_j)$ coincides with a special case of a term used to capture competition in Dawson's geostochastic logistic model, an SPDE model developed to study the combined effects of demographic stochasticity, spatial dispersion and locally finite carrying capacity [@Dawson1978]. In relation to the example fitness function discussed in \S\ref{deterministic}, consider $\kappa(x_i-x_j)=\mathcal{O}_{ij}(x_i-x_j)$.

\paragraph{The model}
In SM \S\ref{diffuse} we combine our treatment of resource competition with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/Current Version/community_dynamics.png",community_plot,width=6.5,height=5,dpi=320)

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/Current Version/on_pl.png",on_pl,device="png",width=3.5,height=3.5,units="in",dpi=350)

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/Current Version/interaction_networks.png",coevs,device="png",width=6,height=6,units="in",dpi=350)

@@

ggsave("~/Gits/branching.brownian.motion.and.spde/Current Version/on_pl.png",on_pl2,device="png",width=3.5,height=3.5,units="in")
```

@@

\newpage
